### 2020-04-18 README.Rmd
Updated README.Rmd and README.md

## Sources of Data

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. This gives a temporal coverage of 30 days and spatial coverage of 50 units. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**]. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  #kable_styling(latex_options = c("striped", "scale_down")) %>%
#kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

@@

The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods. The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient] and now is a popular methodology included in several Econometrics textbook [e.g., @Greene2003econometric]. To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p. 141). From this milestone, a huge researches used this methodology in spatial framework. Rey and Montouri [-@Rey1999us], Fingleton [-@Fingleton2007multi], Lauridsen et al. [-@Lauridsen2010spatiotemporal], Le Gallo and Dall’Erba [-@Legallo2006evaluating], and Lopez et al. [-@Lopez2017spatial] are well-known examples in this line.

@@

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
-->

<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)

@@

The more relevant characteristic of a SUR model is the dependence structure among the vectors of residuals, namely:

@@

Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$. This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time. In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations, i.e., time periods.
-->

@@

Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24). This flexibility can be reduced and it is posible to impose restrinctions considering some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time. In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations, i.e., time periods.

@@

\noindent where....

As is the case of cross-sectional data, it is possible identify spatial autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several Lagrange Multipliers tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, there are alternative several specifications that have been propose to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. The stack expresion for the spatial lag SUR model (SUR-SLM),

@@

\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.
-->

@@

\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).

<!--
This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$). The correspond Wald test is available in the R package `spsur`. 
-->

This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the R package `spsur`.

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).

Another alternative methodologies could be use. By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g. Elhorst 2014, Cap. 4), but those models are more dont take account correlation between errors. Therefore, a spatial SUR approach is more reasonable for our purpose.

@@

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

@@

```{r descriptives-temperature, echo = FALSE, fig.height=7, fig.cap="\\label{fig:descriptives-temperature} Distribution of mean temperatures and humidities in the Autonomous Communities in Spain between March 12, 2020 and April 11, 2020. The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost"}
# Summary of temperature by Autonomous Community
temp.summary <- covid19_spain %>% 
  group_by(CCAA, Date) %>%
  summarize(Value = mean(Mean_Temp)) %>%
  mutate(Variable = "Mean Temperature in Community")

@@

```{r control-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)"}
gdppc.plot <- provinces_spain %>%
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc)) +
  scale_fill_distiller(name = "GDP per capita", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")

@@

```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)"}
covid19_spain %>% 
  group_by(Date) %>%
  summarize(Older = cor(log(Older), log(Incidence)),
            GDPpc = cor(log(GDPpc), log(Incidence)),
            Density = cor(log(Density), log(Incidence)),
            Humidity_lag8 = cor(Humidity_lag8, Incidence),
                        Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
                        Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
  select(-Date) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  mutate(Variable = factor(Variable,
                           levels = c("GDPpc", 
                                      "Older", 
                                      "Density", 
                                      "Humidity_lag8",
                                      "Humidity_lag11",
                                      "Humidity_lag11w",
                                      "Mean_Temp_lag8",
                                      "Mean_Temp_lag11",
                                      "Mean_Temp_lag11w"))) %>%
  ggplot(aes(x = Variable, y = Correlation)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90))
```

@@

```{r data-preparation-for-modelling, include=FALSE}
# Organize data for SUR modelling 
GPanel <- plm::pdata.frame(covid19_spain %>%
                             select(province, 
                                    Date,
                                    Incidence, 
                                    Median_Age,
                                    Male2Female,
                                    Older,
                                    GDPpc,
                                    Density,
                                    Transit,
                                    Mean_Temp_lag8,
                                    Humidity_lag8,
                                    Sunshine_Hours_lag8,
                                    Mean_Temp_lag11,
                                    Humidity_lag11,
                                    Sunshine_Hours_lag11,
                                    Mean_Temp_lag11w,
                                    Humidity_lag11w,
                                    Sunshine_Hours_lag11w),
                           c("province","Date"))

@@

```{r spatial-weights, include=FALSE}
# Create spatial weights matrix: 
Wmat <- provinces_spain %>%
  #drop_na() %>%
  as("Spatial") %>%
  poly2nb(queen = FALSE) %>%
  nb2mat(zero.policy = TRUE)

>>>>

## Data

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**]. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset begins on March 13, which is the first date when every province had reported at least one case of COVID-19, and continues until April 11, for a period of 30 days. The unit of analysis is the province. Provinces are the equivalent of states, and are embedded in Autonomous Communities. As an example, Cataluña is an Autonomous Community and consists of four provinces, namely Barcelona, Gerona, Lerida, and Tarragona. The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}. The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the US) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the US). While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is inconsistent at smaller geographies, or cases are not reported at that level at all. The analysis must therefore be considered ecological.

```{r descriptive-statistics, echo=FALSE, cache=TRUE}
data.frame(Variable = c("COVID-19 Incidence",
                        "Area", 
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "Area of province in sq.km",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                    mean(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                    mean(provinces_spain$GDPpc),
                    mean(provinces_spain$Older),
                    mean(provinces_spain$Density),
                    mean(covid19_spain$Mean_Temp),
                    mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                  sd(provinces_spain$GDPpc),
                  sd(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                  sd(provinces_spain$Older),
                  sd(provinces_spain$Density),
                  sd(covid19_spain$Mean_Temp),
                  sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
  #kable_styling(latex_options = c("striped", "scale_down")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease. Lauer et al. [-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we judge it best to use lagged values of the environmental variables. We test different time lags as follows. We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*). Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*). Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*). The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al. [-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418. With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.

@@

The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consisits of cross-sections for different time periods. The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient], and has become a popular methodology included in several econometrics textbook [e.g., @Greene2003econometric]. To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective. In his landmark text, Anselin discussed a model made of "an equation for each time period, which is estimated for a cross section of spatial units" (p. 141). From this milestone, a large body of research has developed to extend the classical SUR into a spatial framework [e.g., @Rey1999us; @Lauridsen2010spatiotemporal; @Legallo2006evaluating; @Lopez2017spatial].

@@

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals. 
-->

<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->

![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)

@@

A key feature of the SUR model is the dependence structure among the vectors of residuals, namely:

@@

Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$. This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time. In this case, we can reformulate the coefficients expression $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant across equations. This is equivalent to specifying some effects to be invariant over time.
-->

@@

Note that this specification is very flexible, in that it allows changes in the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24). This flexibility can be reduced and it is posible to impose restrictions to consider some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients as being constant over time. In this case, we can reformulate the coefficients expression ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant across equations. This is equivalent to specifying some effects to be invariant over time.

@@

As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio]. While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid]. For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.

The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:

@@

\noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.
-->

@@

\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).

<!--
This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$). The correspond Wald test is available in the `R` package `spsur`.
-->

This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`.

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019). Another alternative methodologies could be use. By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g. Elhorst 2014, Cap. 4), but those models are more dont take account correlation between errors. Therefore, a spatial SUR approach is more reasonable for our purpose.

@@

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE, cache = TRUE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

@@

```{r descriptives-temperature, echo = FALSE, fig.height=7, fig.cap="\\label{fig:descriptives-temperature} Distribution of mean temperatures and humidities in the Autonomous Communities in Spain between March 12, 2020 and April 11, 2020. The Autonomous Communities have been sorted by latitude, with communities to the left being the northermost, and to the right the southernmost", cache = TRUE}
# Summary of temperature by Autonomous Community
temp.summary <- covid19_spain %>% 
  group_by(CCAA, Date) %>%
  summarize(Value = mean(Mean_Temp)) %>%
  mutate(Variable = "Mean Temperature in Community")

@@

```{r control-map, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:control-map}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
gdppc.plot <- provinces_spain %>%
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc)) +
  scale_fill_distiller(name = "GDP per capita", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")

@@

```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)", cache = TRUE}
covid19_spain %>% 
  group_by(Date) %>%
  summarize(Older = cor(log(Older), log(Incidence)),
            GDPpc = cor(log(GDPpc), log(Incidence)),
            Density = cor(log(Density), log(Incidence)),
            Humidity_lag8 = cor(Humidity_lag8, Incidence),
            Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
            Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
  select(-Date) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  mutate(Variable = factor(Variable,
                           levels = c("GDPpc", 
                                      "Older", 
                                      "Density", 
                                      "Humidity_lag8",
                                      "Humidity_lag11",
                                      "Humidity_lag11w",
                                      "Mean_Temp_lag8",
                                      "Mean_Temp_lag11",
                                      "Mean_Temp_lag11w"))) %>%
  ggplot(aes(x = Variable, y = Correlation)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90))
```

@@

```{r data-preparation-for-modelling, include=FALSE, cache = TRUE}
# Organize data for SUR modelling 
GPanel <- plm::pdata.frame(covid19_spain %>%
                             select(province, 
                                    Date,
                                    Incidence, 
                                    Median_Age,
                                    Male2Female,
                                    Older,
                                    GDPpc,
                                    Density,
                                    Transit,
                                    Mean_Temp_lag8,
                                    Humidity_lag8,
                                    Sunshine_Hours_lag8,
                                    Mean_Temp_lag11,
                                    Humidity_lag11,
                                    Sunshine_Hours_lag11,
                                    Mean_Temp_lag11w,
                                    Humidity_lag11w,
                                    Sunshine_Hours_lag11w),
                           c("province","Date"))

@@

```{r spatial-weights, include=FALSE, cache = TRUE}
# Create spatial weights matrix: 
Wmat <- provinces_spain %>%
  #drop_na() %>%
  as("Spatial") %>%
  poly2nb(queen = FALSE) %>%
  nb2mat(zero.policy = TRUE)
